```{r setup}
#install.packages("dplyr")
library(dplyr)
library(ggplot2)
```

We will be practicing with dplyr function using the `starwars` dataset:

```{r}
head(starwars)

```

## Subset rows with `filter`

`filter()` allows us to subset the **rows** of a dataset based on the values

```{r}
## Filter for rows with human Star Wars characters
filter(starwars, species == "Human")

## save to a new R object
humans <- filter(starwars, species == "Human")
humans
filter(humans, sex == "male")

## filter multiple columns at once
filter(starwars, species == "Human", sex == "male")
```

We can use different comparison operators to filter:

-   `!` is "not

-   `==` is "equal"

-   `&` is "and"

-   `|` is "or"

-   `<` is "less than"

-   `>` is "greater than"

-   `is.na()` returns TRUE if value is NA

```{r}
## return characters with a missing mass
filter(starwars, is.na(mass))
```

```{r}
## How to get all character that are not human
filter(starwars, species != "Human")
```

variable_name `%in%` vector_of_values returns TRUE if the value of variable_name is equal to one of the values in vector_of_values

This is a shortcut for variable_name == vector_of_values[0} \| variable_name == vector_of_values[2]...

```{r}
skin_color <- "grey"
skin_color %in% c("yellow", "fair", "tan") #returns false because "grey" is not in vector of colors

## How would we filter for characters with brown or blond hir
filter(starwars, hair_color %in% c("blond", "brown"))
```

grepl("characters_of_interest", variable_name) returns TRUE if the value of variable_name contains the characters_of_interest

```{r}
filter(starwars, grepl("blond", hair_color))

##how would we get all characters with blue in their eyes
filter(starwars, grepl("blue", eye_color))
```

You can use `<`,`>`,`<=`,`>=` to compare numeric variables or categorical variables

-   nominal categorical variables (character vectors) are ranked alphabetically

-   ordinal categorical variables (factors) are ranked based on the built-in order

```{r}
## starwars characters with a name starting with letters before "D"
filter(starwars, name < "D")
```

`between(x, left, right)` is a shorthand for x \>= left & x \<= right

```{r}
between(5, 1, 7) # true because 1<=5<=7
filter(starwars, between(mass, 50, 100))
```

## Practice!

Find all starwars characters that

1.  Are missing a hair color

2.  Are shorter than 100cm

3.  Were born between the years 100 and 200

4.  Another of your choice

```{r}
#1
filter(starwars, is.na(hair_color))

#2
filter(starwars, height < 100, species != "Droid")

#3
filter(starwars, between(birth_year, 100, 200))
```

## Order rows with arrange

`arrange()` reorders rows based on the values of column(s)

```{r}
arrange(starwars, name)
# order of character values: aAbBcCdD

## order by species and birth year
arrange(starwars, species, birth_year)

#use desc() to get the brith years in descending order
arrange(starwars, species, desc(birth_year))
```

Practice!

1.  Arrange the starwars dataset from tallest to shortest characters

2.  Arrange the starwars dataset form shortest to tallest by species

3.  Alphabetize the starwars charcters by they hair color then eye color

```{r}
#1
arrange(starwars, desc(height))

#2
arrange(starwars, height, species)

#3
arrange(starwars, hair_color, eye_color)
```

## Subset columns with `select()`

```{r}
select(starwars, hair_color, skin_color, eye_color)
character_colors <- select(starwars, hair_color, skin_color, eye_color)
character_colors
```

We can also select multiple columns in a row

```{r}
select(starwars, name: eye_color)
```

We can also remove columns by name using a -

```{r}
select(starwars, -films, -vehicles, -starships)
select(starwars, -c(films, vehicles, starships))
select(starwars, -c(films:starships))
```

Helpful functions

`everything()` is a shortcut for "everything else"

```{r}
select(starwars, species, everything()) # put species first, then all the other columns

## alphabetize columns
starwars_names <- sort(names(starwars))
starwars_names
select(starwars, starwars_names)
```

`starts_with("abc")` matches column names that start with "abc"

`ends_with("abc")` matches column names that end with "abc"

`contains("abc")` matches column names that contain "abc"

```{r}
select(starwars, starts_with("h"))

select(starwars, contains("or"))
```

Practice!

1.  Select each character's homeworld and species. What question might this subset of variables help answer?

2.  Brainstorm as many ways as possible to select hair_color, skin_color, eye_color, and birth_year

```{r}
#1
select(starwars, homeworld, species)

#this can answer the question of how concentrated a singular species is on each homeworld

#2
select(starwars, contains("_"))
select(starwars, contains("color"), birth_year)
select(starwars, hair_color:birth_year)
select(starwars, hair_color, skin_color, eye_color, birth_year)
select(starwars, -c(name:mass, sex:starships))
```

## Add new variables with `mutate`

```{r}
starwars_small <- select(starwars, height, mass, species)
starwars_small

mutate(starwars_small,
       height_m = height/100)
```

we can also create multiple variables at once and we can refer to variables we just created

```{r}
mutate(starwars_small,
       height_m = height/100,
       bmi = mass/height_m^2)

## only keep new variables
transmute(starwars_small,
       height_m = height/100,
       bmi = mass/height_m^2)
```

Summary statistics give one value for the entire column: `sum()`, `mean()`

```{r}
mutate(starwars_small,
       sum = sum(mass, na.rm = T))

mutate(starwars_small,
       row_addition = height + mass,
       summary_sum = sum(height, mass, na.rm = T))
```

Group a variable using the `.by` parameter

```{r}
mutate(starwars_small,
       species_mass = mean(mass, na.r = T),
       .by = species)
```

Change the type of a variable

```{r}
mutate(starwars_small,
       mass = as.logical(mass))
```

Practice

1.  convert mass to pounds (1 kg \~ 2.2 lbs)

```{r}
mutate(starwars_small,
       weight_lbs = mass * 2.2)
```

## Summarize data with `summarize`

```{r}
summarize(starwars,
          mean_mass = mean(mass, na.rm = T))

#Group by species
summarize(starwars,
          mean_mass = mean(mass, na.rm = T),
          .by = species)
```

The only columns kept are the columns you group by and the columns your create

`n()` is a helpful function that returns the number of rows in a group

```{r}
summarize(starwars,
          mean_mass = mean(mass, na.rm = T),
          count = n(),
          .by = species)

## group by multiple groups
summarize(starwars,
          mean_mass = mean(mass, na.rm = T),
          count = n(),
          .by = c(species, sex))
```
